Type Refinements for the Working Class

### 2015-11-15 paper.tex
Simplify semantic universe (close #4); show how to externalize internal judgments
- change semantic universe to Set^Sctx for simplicity (#4)

- changed \Omega to \Theta for metavariable contexts to avoid confusion
  w/ subobject classifier

- improve the definition of internal judgments, and show how to
  externalize them

- comment out some of the computation system stuff for now until I get
  things in order

\newcommand\Hom[3]{#1\!\left[#2,\,#3\right]\!}

@@

\newcommand\SemBrackets[1]{\left\llbracket#1\right\rrbracket}
\newcommand\Yoneda[1]{\boldsymbol{y} (#1)}

@@

For any abt signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we can deploy
a generic logical framework. Recall from~\cite{sterling-morrison:2015} that
$\SCtx$ is the category of symbol-contexts and their injective, sort-preserving
renamings. Then, fix a small category $\mathcal{W}$ of worlds along with a map
$\Of{\pi}{\mathcal{W}\ra\SCtx}$ such that $\pi$ is a Grothendieck
fibration~\cite{jacobs:2001, taylor:1999}:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&, row sep = large]
    {\IMode{\mathcal{W}}}
      \arrow[dash, d]
\&
    {\IMode{{w}_{i}}}
      \arrow[drrr]
      \arrow[dotted, dr]
      \arrow[mapsto, dd]
\&
    \null
\&
    \null
\&
    \null
\\
    {\IMode{{\pi}}}
      \arrow[d]
\&
    \null
\&
    {\OMode{{w}_{j}}}
      \arrow[rr, swap, "\OMode{{\iota}_{\varrho}}"]
      \arrow[mapsto, dd, bend right = 50]
      \arrow[mapsto, d]
\&
    \null
\&
    |[alias = Wk]|{\IMode{{w}_{k}}}
      \arrow[mapsto, d]
\\
    {\IMode{\SCtx}}
\&
    |[alias = Ui]|{\IMode{{\Upsilon}_{i}}}
      \arrow[dr]
\&
    {\OMode{\scriptstyle{{\pi}({w}_{j})}}}
      \arrow[rr, "\OMode{{\pi}({\iota}_{\varrho})}", ""{name = Msrc, swap}]
      \arrow[equals, d]
\&
    \null
\&
    {\IMode{\scriptstyle{{\pi}({w}_{k})}}}
      \arrow[equals, d]
\\
    \null
\&
    \null
\&
    {\IMode{{\Upsilon}_{j}}}
      \arrow[rr, swap, "{\IMode{\varrho}}", ""{name = Mtgt, swap}]
\&
    \null
\&
    |[alias = Uk]|{\IMode{{\Upsilon}_{k}}}
      \arrow[mapsto, from = Wk, to = Uk, bend right = 50, crossing over]
      \arrow[equals, from = Msrc, to = Mtgt]
      \arrow[from = Ui, to = Uk, crossing over]
  \end{tikzcd}
\]
Then, a \emph{judgment} shall be interpreted as an object in the presheaf
category $\Define{\mathbb{J}}{\Sets^{\OpCat{\mathcal{W}}}}$; in other words, a
judgment is identified with the intensional set of its renaming-sensitive
derivations.

\begin{definition}[Hypothetical Judgment]
  For two judgments $\mathcal{J}_1$ and $\mathcal{J}_2$, we can form the
  hypothetical judgment $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ using the
  exponential construction~\cite[p.~46]{maclane-moerdijk:1992}, which expresses the semantic consequence
  of $\mathcal{J}_2$ from $\mathcal{J}_1$:
  \begin{align*}
    \ADefine{
      \left(\HypJ{\mathcal{J}_2}{\mathcal{J}_1}\right)(w)
    }{
      \left({\mathcal{J}_2}^{\mathcal{J}_1}\right)(w)
    }\\
    &\cong
    \OMode{
      \Hom{\mathbb{J}}{
        \Yoneda{w}\times\mathcal{J}_1
      }{
        \mathcal{J}_2
      }
    }\\
    &\cong
    \OMode{
      \int_{w'}{{{w'}\succeq{w}}\Rightarrow{{{\mathcal{J}}_{2}(w')}^{{\mathcal{J}}_{1}(w')}}}
    }
  \end{align*}

  In other words, the hypothetical judgment
  $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ is evident when there shall forevermore
  be an effective transformation of evidence for the antecedent into evidence
  for the consequent. It is important to keep in mind that this is a
  \emph{semantic} consequence and therefore expresses admissibilities, in
  contrast to the \emph{logical} consequence which expresses
  derivabilities~\cite{harper:2016}. We will not have any need for the latter.
\end{definition}

\begin{definition}[General Judgment]
  For a metavariable context $\IsMetaCtx{\Omega}$, we can define the set
  of its substitutions as follows:
  \[
    \Define{{\square\Omega}}{
      \prod_{(\MV{m}:v)\in\Omega}
        \MkSet{E\mid\IsAbtUnmoded{\cdot}{\cdot}{\cdot}{E}{v}}
    }
  \]

  For any diagram $\Of{\mathcal{J}}{\mathbb{J}^{\square\Omega}}$, we can define
  the general judgment $\GenJ{\Omega}{\mathcal{J}}$ as follows, where
  $\Of{\Yoneda{-}}{\mathcal{W}\ra\mathbb{J}}$ is the Yoneda Embedding:
  \begin{align*}
    \ADefine{
      \left(\GenJ{\Omega}{\mathcal{J}}\right)(w)
    }{
      \Hom{\mathbb{J}}{
        \Yoneda{w}
      }{
        \varprojlim\mathcal{J}
      }
    }\\
    &\cong
    \OMode{
      \int_{(w',\,\vec{E})}{{w'}\succeq{w}\Rightarrow{\mathcal{J}(\vec{E})(w')}}
    }
  \end{align*}

   Intuitively, the general judgment $\GenJ{\Omega}{\mathcal{J}}$ is evident
   when the family of judgments $\mathcal{J}$ shall forevermore have a global
   section (i.e.\ for any substitution $\Member{\vec{E}}{\square\Omega}$,
   $\mathcal{J}(\vec{E})$ shall be evident). Recall that metavariables have
   \emph{valences} rather than sorts; as such, the general judgment allows us
   to quantify over terms of higher type.

@@

   We will write $\HPGenJ{\Upsilon}{\Gamma}{\Omega}{\mathcal{J}}$ for
   $\GenJ{\Omega'}{\mathcal{J}}$, where
   $\Lookup{\Omega'}{\MV{m}}{\MkValence{\Upsilon,\vec{\sigma}}{\Gamma,\vec{\tau}}{\tau}}$
   if $\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$.
   Additionally, for clarity we will often write
   $\GenJ{\MV{m}:\sigma}{\mathcal{J}}$ rather than
   $\GenJ{\MV{m}:\MkValence{\cdot}{\cdot}{\sigma}}{\mathcal{J}}$; likewise, in
   the body of $\mathcal{J}$, when there are contextually salient
   $\vec{u}$, $\vec{x}$, we will simply write $\MV{m}$ rather than the more proper
   $\MApp{\MV{m}}{\vec{u}}{\vec{x}}$.

@@

@@

In this section we generalize Howe's notion of lazy computation
system~\cite{howe:1989} to the multi-sorted, symbol-parameterized setting. An
\emph{lazy computation language} is an abt signature
$\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$ along with a distinguished copresheaf
$\Of{\CanOperators}{\Sets^{\SCtx\times\Discrete{\Arities}}}$ of
\emph{canonical} operators such that $\IsSubsetEq{\CanOperators}{\Operators}$.
For a lazy computation language
$\Match{L}{\Tuple{\Sorts,\Operators,\CanOperators}}$, let us define the
copresheaves of closed expressions and closed values over symbol contexts and sorts as
follows:
\begin{align*}
  \ADefine{\Exprs_\tau(\Upsilon)}{
    \MkSet{M \mid \IsAbtUnmoded{\cdot}{\Upsilon}{\cdot}{M}{\tau}}
  }\\
  \ADefine{\Values_\tau(\Upsilon)}{
    \MkSet{
      M\equiv\App{\vartheta}{\vec{E}}
      \mid M\in\Exprs_\tau(\Upsilon)\land \exists a. \CanOperators\Pair{\Upsilon}{a}\ni\vartheta
    }
  }
\end{align*}

Then, a \emph{lazy computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with an indexed
$\equiv_\alpha$-functional relation
$\Of{\Downarrow}{\Pow{\Exprs\times\Values\times\NaturalsPsh}}$ where
$\NaturalsPsh$ is the constant presheaf $\MapsTo{(\Upsilon, \tau)}{\Naturals}$.
We will write $\EvalN{\Upsilon}{\tau}{n}{M}{N}$ when
$\Member{\Tuple{M,N,n}}{\Downarrow_\tau(\Upsilon)}$, to mean that $M$ evaluates
to $N$ in $n$ steps.

>>>>

\newcommand\Hom[3]{#1\!\left[#2,\,#3\right]}

@@

\newcommand\SemBrackets[1]{\left\llbracket#1\right\rrbracket}
\newcommand\Embed[1]{\left\lfloor#1\right\rfloor}
\newcommand\Yoneda[1]{\boldsymbol{y} (#1)}

@@

For any abt signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we can deploy
a generic logical framework. Recall from~\cite{sterling-morrison:2015} that
$\SCtx$ is the category of symbol-contexts and their injective, sort-preserving
renamings. Then, a \emph{judgment} shall be interpreted as an object in the presheaf
category $\Define{\mathbb{J}}{\Sets^\SCtx}$; in other words, a
judgment is identified with the intensional set of its renaming-sensitive
derivations. $\mathbb{J}$, like all presheaf categories, is a topos.

\begin{definition}[Hypothetical Judgment]
  For two judgments $\mathcal{J}_1$ and $\mathcal{J}_2$, we can form the
  hypothetical judgment $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ using the
  exponential construction~\cite[p.~46]{maclane-moerdijk:1992}, which expresses
  the semantic consequence of $\mathcal{J}_2$ from $\mathcal{J}_1$:
  \begin{align*}
    \ADefine{
      \left(\HypJ{\mathcal{J}_2}{\mathcal{J}_1}\right)(\Upsilon)
    }{
      \left({\mathcal{J}_2}^{\mathcal{J}_1}\right)(\Upsilon)
    }\\
    &\cong
    \OMode{
      \Hom{\mathbb{J}}{
        \Yoneda{\Upsilon}\times\mathcal{J}_1
      }{
        \mathcal{J}_2
      }
    }\\
    &\cong
    \OMode{
      \int_{\Upsilon'}{{\Upsilon\inj\Upsilon'}
        \Rightarrow{{{\mathcal{J}}_{2}(\Upsilon')}^{{\mathcal{J}}_{1}(\Upsilon')}}}
    }
  \end{align*}

  In other words, the hypothetical judgment
  $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ is evident when there shall forevermore
  be an effective transformation of evidence for the antecedent into evidence
  for the consequent. It is important to keep in mind that this is a
  \emph{semantic} consequence and therefore expresses admissibilities, in
  contrast to the \emph{logical} consequence which expresses
  derivabilities~\cite{harper:2016}. We will not have any need for the latter.

  \paragraph{Notation}
  For a hypothetical judgment with multiple hypotheses, we will write
  $\HypJ{\mathcal{J}_3}{\mathcal{J}_1, \mathcal{J}_2}$ rather than
  $\HypJ{\mathcal{J}_3}{\mathcal{J}_1\times\mathcal{J}_2}$ or
  $\HypJ{(\HypJ{\mathcal{J}_3}{\mathcal{J}_2})}{\mathcal{J}_1}$.
\end{definition}

\begin{definition}[General Judgment]
  For a metavariable context $\IsMetaCtx{\Theta}$, we can define the set
  of its substitutions as follows:
  \[
    \Define{{\square\Theta}}{
      \prod_{(\MV{m}:v)\in\Theta}
        \MkSet{E\mid\IsAbtUnmoded{\cdot}{\cdot}{\cdot}{E}{v}}
    }
  \]

  For any diagram $\Of{\mathcal{J}}{\mathbb{J}^{\square\Theta}}$, we can define
  the general judgment $\GenJ{\Theta}{\mathcal{J}}$ as follows, where
  $\Of{\Yoneda{-}}{\OpCat{\SCtx}\ra\mathbb{J}}$ is the Yoneda Embedding:
  \begin{align*}
    \ADefine{
      \left(\GenJ{\Theta}{\mathcal{J}}\right)(\Upsilon)
    }{
      \Hom{\mathbb{J}}{
        \Yoneda{\Upsilon}
      }{
        \varprojlim\mathcal{J}
      }
    }\\
    &\cong
    \OMode{
      \int_{(\Upsilon',\,\vec{E})}{\Upsilon\inj\Upsilon'\Rightarrow{\mathcal{J}(\vec{E})(\Upsilon')}}
    }
  \end{align*}

   Intuitively, the general judgment $\GenJ{\Theta}{\mathcal{J}}$ is evident
   when the family of judgments $\mathcal{J}$ shall forevermore have a global
   section (i.e.\ for any substitution $\Member{\vec{E}}{\square\Theta}$,
   $\mathcal{J}(\vec{E})$ shall be evident). Recall that metavariables have
   \emph{valences} rather than sorts; as such, the general judgment allows us
   to quantify over terms of higher type.

@@

   We will write $\HPGenJ{\Upsilon}{\Gamma}{\Theta}{\mathcal{J}}$ for
   $\GenJ{\Theta'}{\mathcal{J}}$, where
   $\Lookup{\Theta'}{\MV{m}}{\MkValence{\Upsilon,\vec{\sigma}}{\Gamma,\vec{\tau}}{\tau}}$
   if $\Lookup{\Theta}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$.
   Additionally, for clarity we will often write
   $\GenJ{\MV{m}:\sigma}{\mathcal{J}}$ rather than
   $\GenJ{\MV{m}:\MkValence{\cdot}{\cdot}{\sigma}}{\mathcal{J}}$; likewise, in
   the body of $\mathcal{J}$, when there are contextually salient
   $\vec{u}$, $\vec{x}$, we will simply write $\MV{m}$ rather than the more proper
   $\MApp{\MV{m}}{\vec{u}}{\vec{x}}$.

@@

\begin{definition}[Internal Judgments]

  It often happens that the evidence for a judgment is the knowledge of the
  meaning of a new form of judgment. This can be neatly expressed in our topos
  using the subobject classifier $\Omega$, where the evidence of $\Omega$ at
  $\Upsilon$ gives rise to an \emph{internal judgment} which is defined at all
  $\Upsilon'$ for which we have $\Of{\rho}{\Upsilon\inj\Upsilon'}$.  A
  \emph{global internal judgment} is a section
  $\Of{J}{\int_\Upsilon\Hom{\mathbb{J}}{\Yoneda{\Upsilon}}{\Omega}}$, which can
  be externalized as a judgment $\Of{\Embed{J}}{\mathbb{J}}$ whose derivations
  at $\Upsilon$ are the arrows
  $\Of{\phi}{\Hom{\mathbb{J}}{\Yoneda{\Upsilon}}{\mathbbold{1}}}$ such that the
  following diagram commutes:
  \[
    \begin{tikzcd}
      \IMode{\Yoneda{\Upsilon}}
        \arrow[r, dashed, "\OMode{\phi}"]
        \arrow[rd, "\IMode{J_\Upsilon}"']
      & \IMode{\mathbbold{1}} \arrow[d, "\IMode{\mathbf{true}}"] \\
      & \IMode{\Omega}
    \end{tikzcd}
  \]

  In most cases, we will say something like ``To know $\mathcal{J}$ is to have
  a judgment $\mathcal{J'}$'' rather than appealing directly to the subobject
  classifier.
\end{definition}

@@

In this section we generalize Howe's notion of lazy computation
system~\cite{howe:1989} to the multi-sorted, symbol-parameterized setting. An
\emph{lazy computation language} is an abt signature
$\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$ along with a distinguished copresheaf
$\Of{\CanOperators}{\Sets^{\SCtx\times\Discrete{\Arities}}}$ of
\emph{canonical} operators such that $\IsSubsetEq{\CanOperators}{\Operators}$.
For a lazy computation language
$\Match{L}{\Tuple{\Sorts,\Operators,\CanOperators}}$, let us define the
presheaves of closed expressions and closed values as follows:
\begin{align*}
  \ADefine{\Exprs_\tau(\Upsilon)}{
    \MkSet{M \mid \IsAbtUnmoded{\cdot}{\Upsilon}{\cdot}{M}{\tau}}
  }\\
  \ADefine{\Values_\tau(\Upsilon)}{
    \MkSet{
      M\equiv\App{\vartheta}{\vec{E}}
      \mid M\in\Exprs_\tau(\Upsilon)\land \exists a. \CanOperators\Pair{\Upsilon}{a}\ni\vartheta
    }
  }
\end{align*}

\begin{remark}
  If read as a judgment, $\Exprs_\tau$ (resp.\ $\Values_\tau$) means ``there is
  a closed expression (resp.\ value) of sort $\tau$'', and has as its evidence any closed
  expression (resp.\ value) of sort $\tau$.
\end{remark}

Then, a \emph{lazy computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with an $\SCtx$-indexed
$\equiv_\alpha$-functional evaluation relation
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$ presupposing $\Member{n}{\Naturals}$,
$\Member{M}{\Exprs_\tau(\Upsilon)}$ and $\Member{N}{\Values_\tau(\Upsilon)}$,
expressing that $M$ evaluates to $N$ in $n$ steps.

%% \begin{remark}
%%
%%   Formally, the evaluation relation is given in our logical framework by a
%%   presheaf of evaluation rules $\Of{\mathfrak{R}^\tau_\Downarrow}{\mathbb{J}}$,
%%   such that the judgment
%%   $\HypJ{\Pow{\mathfrak{R}^\tau_\Downarrow}}{\Exprs_\tau, \Values_\tau,
%%   \NaturalsPsh}$ is globally evident where $\NaturalsPsh$ is the constant
%%   presheaf $\MapsTo{\Upsilon}{\Naturals}$. In other words, we have a family of
%%   internal judgments $\Of{[\Downarrow_\tau](\Upsilon)}{\int_{\Upsilon'}
%%   \Upsilon\inj\Upsilon'\Rightarrow\Pow{\mathfrak{R}^\tau_\Downarrow}}$
%%
%%
%% shall be evident by means of some internal
%% family of judgments $[\Downarrow_\tau]$, where $\NaturalsPsh$ is the constant
%% presheaf $\MapsTo{\Upsilon}{\Naturals}$. In other words, we shall forevermore know
%% what it means to know that some expression evaluates to some value in $n$
%% steps.
%% \end{remark}
%%
%% We will write $\Eval{\Upsilon}{\tau}{M}{N}$ to mean that there exists an $n$
%% such that $\EvalN{\Upsilon}{\tau}{n}{M}{N}$; in other words, the judgment is
%% interpreted as the following presheaf:
%% \[
%%   \Define{
%%     \left(\Upsilon\parallel M\Downarrow_\tau N\right)(w)
%%   }{
%%     \int^n[\Downarrow_\tau](w,M,N,n)
%%   }
%% \]
%%
%%
%% By this token, the derivations of $\EvalN{\Upsilon}{\tau}{n}{M}{N}$ at
%% $w$ are the set $[\Downarrow_\tau](w,M,N,n)$; we write
%% $\Eval{\Upsilon}{\tau}{M}{N}$ for the judgment
%% $\int^n[\Downarrow_\tau](-,M,N,n)$
%
